Appendix S1. Linear regression models and non-linear population dynamics

### 2024-11-22 manuscript/Appendix_S1_v02.Rmd
Cleaning up Appendix S1

---
title: "Appendix S1. Linear regression models and non-linear population dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
library(MASS)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

## Population dynamics simulation
In the metacommunity simulation in the main text, a species resides in a site with an initial population size $N_{i,0} \sim Pois(10)$, a growth rate $r_i$ that depends on the local environmental value $E_k$ and the species trait $x_i$, and a fixed intra-specific competition coefficient of $\alpha_{ii}$ = 0.00125. We simulate population growth here:

```{r}
set.seed(42)
# Simulate initial species population growth
N1.0 <- rpois(1,10)
r1.0 <- 1.67
alpha.11 <- 0.00125
# model function
disc_log <- function(r, N0, alpha) {
    Nt1 <- (r*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 30
N <- rep(NA, t)
N[1] <- N1.0
for (i in 2:t) {
  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
}
```
```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
# Plot simulation: ggplot
dat <- as.data.frame(N)
dat$time <- 1:t
ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
    linetype = "dashed", color = "gray")
```

@@

$$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
```{r,warning=FALSE}
# Fit the model
m.1.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS")
m.1.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]))
#plotting the series along with the fitted values
m.1.ar.fit <- log(N) - residuals(m.1.ar)
m.1.lm.fit <- log(dat$N[2:t]) - m.1.lm$resid
dat$ar1.fit <- m.1.ar.fit
dat$lm.fit <- NA
dat$lm.fit[2:t] <- m.1.lm.fit
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

@@

```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
dN <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN[i] <- log(N[i+1])-log(N[i])
}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
```

@@

```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Change in population size from one time step to the next $N_{t+1}$ as a function of $N_{t+1}$"}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
```
## Bayesian linear statistical model: HMSC
We can estimate the same model parameters using HMSC:

```{r}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- data.frame(x=log(dat$N[1:(t-1)]))
m.1.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
#preds <-computePredictedValues(m.1.sample)
plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
```

```{r}
lm_dat <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)])))
colnames(lm_dat) <- c("Nt1","Nt")
ggplot(lm_dat,aes(Nt,Nt1)) +
  stat_summary(fun.data= mean_cl_normal) + 
  geom_smooth(method='lm')
```

@@

Here, $\hat{W}$ is the maximal population growth rate (set to `r r1.0` as above), $E$ is the local environmental trait optimum value, and $x_{i,t}$ is species *i* trait value at time *t*. We see that if $E = x_{i,t}$ then the growth rate is at the value $r$ = `r r1.0`. Here, we begin with $E = x_{i,t} = 0.8$, then simulate the environment $E$ value fluctuating randomly over time, and finally use a linear model to fit $E$ as a covariate.

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
E.0 <- 0.8
x1.0 <- 0.8
# model function
disc_log_E <- function(r, N0, alpha, E, x) {
    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 40
N <- rep(NA, t)
N[1] <- N1.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
  E[i] <- E[i-1] + rnorm(1,0,.1)
}
```
```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. Relationship between E and Nt is also shown."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N,E))
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
p1 + p2
```

@@

```{r,warning=FALSE}
# Fit the model
m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + log(E[1:(t-1)]))
#plotting the series along with the fitted values
m.2.ar.fit <- log(N) - residuals(m.2.ar)
m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
dat$ar2.fit <- m.2.ar.fit
dat$lm2.fit <- NA
dat$lm2.fit[2:t] <- m.2.lm.fit
```
```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
dN <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN[i] <- log(N[i+1])-log(N[i])
}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()

d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
d.dN$E <- log(dat$E[2:t])
p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()

@@

The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ can still be captured by a linear relationship. However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear. This tells us that the lm is good for predictions, but not for inference (for capturing well the relationsip between the predictor and response variable). The use of linear relationships in JSDMs is discussed in [@Ingram2020], and in many applications (e.g. [@Erickson2023]) quadratic terms are used, which create bell-shaped response curves that may better match species with optimal niches (as opposed to linear, monotonically increasing relationships between population size and environmental predictors). We thus include a quadratic term for $E_t$ to provide a better fit to the data.

@@

```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
ggplot(data = df, aes(x = E, y = Nt1)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
```

@@

```{r}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- df
m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

@@

```{r figs10, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
# Esq is manually constructed as gradient-produced E^2
Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
```

@@

```{r}
# Simulate initial species population growth with a second species present
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.67
r2.0 <- 1.7
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```
## Linear statistical model with covariate for both species

@@

```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N$N1,N$N2))
colnames(dat) <- c("N1","N2")
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point()
p1 + p2
```

>>>>

---
title: "Appendix S1. Linear regression models and non-linear population dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs_S1.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
library(MASS)
library(ecoevor)
library(Hmisc)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

## Population dynamics simulation
We simulate population growth using $r_i = 1.67$ and $\alpha_{ii}$ = 0.00125, with an initial population size $N_{1,0} = 10$. This is implemented using the `disc_log` function from the `ecoevor` package [https://github.com/jhpantel/ecoevoR](https://github.com/jhpantel/ecoevoR).

```{r}
set.seed(42)
# Simulate initial species population growth
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
# Simulation of model for t time steps
t <- 30
N <- rep(NA, t)
N[1] <- N1.0
for (i in 2:t) {
  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
}
```
```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
# Plot simulation: ggplot
dat <- as.data.frame(N)
dat$time <- 1:t
ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
    linetype = "dashed", color = "gray") +
  theme_classic()
```

@@

$$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
```{r,warning=FALSE}
# Fit the model
m.1.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS")
m.1.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]))
#plotting the series along with the fitted values
m.1.ar.fit <- log(N) - residuals(m.1.ar)
m.1.lm.fit <- log(dat$N[2:t]) - m.1.lm$resid
dat$ar1.fit <- m.1.ar.fit
dat$lm.fit <- NA
dat$lm.fit[2:t] <- m.1.lm.fit
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive (grey) and linear regression (orange) model.",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

@@

```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
dN <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN[i] <- log(N[i+1])-log(N[i])
}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point() + theme_classic()
```

@@

```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Change in population size from one time step to the next $\\Delta N$ as a function of $N_{t+1}$"}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point() + theme_classic()
```
## Bayesian linear statistical model: HMSC
We can estimate the same model parameters using HMSC:

```{r results='hide', message=FALSE, warning=FALSE}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- data.frame(x=log(dat$N[1:(t-1)]))
m.1.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis) using HMSC.",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
```

```{r figs6, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs6} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis) using lm.",results='hide',fig.keep='all',warning=FALSE,message=FALSE}
lm_dat <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)])))
colnames(lm_dat) <- c("Nt1","Nt")
ggplot(lm_dat,aes(Nt,Nt1)) +
  stat_summary(fun.data= mean_cl_normal) + 
  geom_smooth(method='lm') + theme_classic()
```

@@

Here, $\hat{W}$ is the maximal population growth rate (set to `r r1.0` as above), $E$ is the local environmental trait optimum value, and $x_{i,t}$ is species *i* trait value at time *t*. We see that if $E = x_{i,t}$ then the growth rate is at the value $r$ = `r r1.0`. Here, we begin with $E = x_{i,t} = 0.8$, then simulate the environment $E$ value fluctuating randomly over time, and finally use a linear model to fit $E$ as a covariate. We implement the simulation using the `disc_log_E` function from the `ecoevor` package.

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
E.0 <- 0.8
x1.0 <- 0.8
# Simulation of model for t time steps
t <- 40
N <- rep(NA, t)
N[1] <- N1.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
  E[i] <- E[i-1] + rnorm(1,0,.1)
}
```
```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs7}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. The value of E over time is also shown."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N,E))
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point() + theme_classic()
p1 + p2
```

@@

```{r,warning=FALSE}
# Fit the model
m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + log(E[1:(t-1)]))
#plotting the series along with the fitted values
m.2.ar.fit <- log(N) - residuals(m.2.ar)
m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
dat$ar2.fit <- m.2.ar.fit
dat$lm2.fit <- NA
dat$lm2.fit[2:t] <- m.2.lm.fit
```

```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs8} Population size over time (black line) with fitted values from a first-order autoregressive model (gray) and a linear model (orange).",warning=FALSE}

colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ can still be captured by a linear relationship. However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear.

```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs9}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. Also shown are change in population size $\\Delta N$ with population size and E."}
dN <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN[i] <- log(N[i+1])-log(N[i])
}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point() + theme_classic()

d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
d.dN$E <- log(dat$E[2:t])
p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point() + theme_classic()
p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point() + theme_classic()
p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point() + theme_classic()

@@

This tells us that the lm is good for predictions, but not for inference (for capturing well the relationsip between the predictor and response variable). The use of linear relationships in JSDMs is discussed in [@Ingram2020], and in many applications (e.g. [@Erickson2023]) quadratic terms are used, which create bell-shaped response curves that may better match species with optimal niches (as opposed to linear, monotonically increasing relationships between population size and environmental predictors). We thus include a quadratic term for $E_t$ to provide a better fit to the data.

@@

```{r figs10, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs10} Population size (logarithm) at one time step $N_{t+1}$ as a function of environment E. Linear and quadratic fits are shown."}
jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
ggplot(data = df, aes(x = E, y = Nt1)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) + theme_classic()
```

@@

```{r results='hide', warning=FALSE, message=FALSE}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- df
m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

```{r}
knitr::knit_exit()
```

@@

```{r figs11, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs11} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
# Esq is manually constructed as gradient-produced E^2
Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
```

@@

```{r}
# Simulate initial species population growth with a second species present
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.67
r2.0 <- 1.7
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs12}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```
## Linear statistical model with covariate for both species

@@

```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs13}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N$N1,N$N2))
colnames(dat) <- c("N1","N2")
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point()
p1 + p2
```
